Modelling single-cell dynamics\\with trajectories and gene regulatory networks

### 2019-11-20 rebuttal/rebuttal.tex

\begin{document}

\section{Summary of most important comments}

\textbf{Overall:} The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis. Certain unpublished chapters feel rushed and unfinished. Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.

\textbf{Overall:} From three jury members: Merge chapters 8-9-10 into one discussion, or merge chapter 8 into chapter 3.

@@

\textbf{dyngen:} {How crucial are certain parts of the GRN model, e.g. mRNA production and degradation?}

\textbf{dyngen:} {Compare against other scRNA-seq generators}

\textbf{dyngen:} {It is claimed dyngen is a multi-modality simulator, yet it produces only transcriptomics data}

\textbf{dyngen:} Chapter does not seem completely finished; may need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like. No references to user documentation, source code (open source?). No discussion about the performance of the framework. Since this is a PhD in
computer science, I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter.

@@

\textbf{dynbenchmark:} {No statistical tests were performed to assess whether methods are significantly different from one another}

\textbf{dynbenchmark:} Turn into UK English to be consistent with the rest of the dissertation.

\textbf{dyno:} {Does not really read as a scientific paper (yet).}

\textbf{dyno:} Chapter does not seem completely finished; especially the results section (4.2) read more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.

@@

\textbf{SCORPIUS:} {The methodology has changed between the bioRxiv preprint and this chapter, yet Figure 5.4 is exactly the same. Did the changes have no (measurable) effect on this analysis?}

\textbf{SCORPIUS:} {Discuss the computational complexity of SCORPIUS.}

\textbf{SCORPIUS:} How objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?

@@

\textbf{bred:} {Better explain how bred is different from GENIE3.}

\textbf{bred:} {What makes bred a true case-wise NI method? Provide a more formal definition of what a case-wise regulatory network is.}

\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + r	egulon specificity score.}

\textbf{bred:} {Discussion is very short, please elaborate. Paper needs finalisation to make it publication-ready.

@@

\textbf{IncGraph:} {On optimising GRNs, compare a simple MST with IncGraph}

\textbf{Benchmarking guidelines:} {As the candidate is not first author on this paper, I’m not sure if this paper
	should be included in the PhD book verbatim.}

\textbf{Discussion:} {The Discussion chapter is missing a more broad context
			description. How are other researchers using these tools to gain new biological insights? Where is the field of modelling single-cell dynamics going?}

\textbf{Discussion:} {Reads more like a summary. Discuss the links between different chapters.}

@@

\subsection{Chapter 10}

\exam{\nimp{The chapter contains a concise recap and critical view on the contributions in
		literature.} It however does not discuss Chapter 9 (see comment above).}

\exam{\imp{The discussion could be improved by stating how the tools are allowing new
		biological insights.} Only a small hint is provided in the last sentence of 10.4.}

\section{Peter Dawyndt}

\exam{Not all chapters have been published and some are still in stage where some work needs to be done before they are publication-ready}

\exam{Not always a logical connection between the successive chapters}

% TODO: provide visual schematic of what topic this dissertation tries to tackle

\exam{Drop the paper X prefix from most of the chapters}

\exam[\tagimp \tagtime]{chapter does not seem completely finished; \imp{may need a working example to demonstrate the steps
	a users needs to take to use the software, and to illustrate how the results look like}; for now, only
	the inner workings of the simulator are discussed}

\exam{no information about implementation language
	(since Gillespie SSA uses a C++ library, this may hint the framework is also implemented in C++)}

\exam{case
	study may also serve as user documentation}

\exam{no references to user documentation, source code
	(open source?)}

\exam{no discussion about the performance of the framework}

\exam[\tagtime]{since this is a PhD in
	computer science, I might also have expected a class diagram on the software, as a proof that the
	framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter}

\exam{p24-25 how are networks specified by user (ergonomy \& standards); support for SBML?}

\exam{p26: has the BBL-ontology been invented for this application or an existing ontology?}

\exam{maybe this chapter should come after the next chapter}

\exam{what reactions have the authors received from the developers of the tools that were reviewed in
	this benchmark, or from users of the TI tools that have used the benchmark results to decide what
	tools to use}

\exam{could this study also yield some standard ways for data formats that tools could use to input/output
	data (already hinted in the text); what approach would the authors choose towards a standards
	proposal?}

\exam{Some users filed an issue on GitHub reporting problems with running Docker on Windows; a
	promise was made by the authors that this issue would be looked at, but no solution has been
	provided so far; any progress in resolving this issue?}

\exam{as with CH2, this chapter does not seem completely finished; especially the results section (4.2) read more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study, e.g. how to run the analysis one the command line, might benefit the readership}

\exam[\tagimp]{fundamental question: \imp{how objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?}}

\exam{paper needs finalisation to make it publication-ready}

\exam{performance measurements show that method is faster than state of the art, but no formal time complexity is derived and no formal proof is given that the method still computes the correct number of graphlets; strange behaviour in time measurements is not diagnosed in depth}

\exam{I would argue against having this as a separate chapter, but would use its content to supplement the introduction and CH3}

\subsection{Discussion}

\exam{points of discussion, open source: upon publication or release-early-release-often?}

\exam[\tagimp]{\imp{reads more like a summary} + listing some use case that made use of contributions made in the manuscript, rather than a real discussion}

\exam{10.7 might better go into a preface of the manuscript}

### 2019-11-21 rebuttal/rebuttal.tex

\textbf{dyngen:} {How crucial are certain parts of the GRN model, e.g. mRNA production and degradation?}

\textbf{dyngen:} {Compare against other scRNA-seq generators}

\textbf{dyngen:} {It is claimed dyngen is a multi-modality simulator, yet it produces only transcriptomics data}

\textbf{dyngen:} May need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like. No references to user documentation, source code (open source?). No discussion about the performance of the framework. 
I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter.

@@

\textbf{dynbenchmark:} {Comparisons focus mainly on relative differences between methods. It would have been interesting to also assess the absolute performance of each method.}

\textbf{dynbenchmark:} {No statistical tests were performed to assess whether methods are significantly different from one another}

\textbf{dynbenchmark:} Turn into UK English to be consistent with the rest of the dissertation.

\textbf{dyno:} Does not really read as a scientific paper (yet). The results section should be improved, it currently reads more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.

@@

\textbf{bred:} {Better explain how bred is different from GENIE3.}

\textbf{bred:} {What makes bred a true case-wise NI method? Provide a more formal definition of what a case-wise regulatory network is.}

\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + r	egulon specificity score.}

\textbf{bred:} {Discussion is very short, please elaborate.}

### 2019-11-23 defence_internal/pres_v1.tex
wip persentation

\newcommand{\summary}[5]{%
	\begin{frame}
		\begin{itemize}
			\item {\color{#1}Single-cell biology}
			\item {\color{#2}Trajectory inference}
			\item {\color{#3}Network inference}
			\item {\color{#4}Benchmarking}
			\item {\color{#5}Practical implications}
		\end{itemize}
	\end{frame}
}

@@

\begin{frame}
	\frametitle{Single-cell biology \uncover<2->{is to study an organism}\\\uncover<3->{by studying the behaviour of its cells}}
	\begin{center}
		\only<2>{\includegraphics[width=\linewidth]{figures/1a_singlecellbiology}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/1b_singlecellbiology}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/1c_singlecellbiology}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Technological advances allow high-throughput\\single-cell biology}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/2_part_scaling.pdf}\\
		\flushright 		{\vspace*{-1em}\footnotesize Regev et al. 2018}
	\end{center}
  \vspace*{-2.2em}
  \uncover<2->{
	  \rotatebox[origin=c]{180}{$\Lsh$} Trajectory inference \\
	  \rotatebox[origin=c]{180}{$\Lsh$} Network inference \\
	}
\end{frame}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{black}{lightgray}{lightgray}{lightgray}

@@

\begin{frame}
	\frametitle{Benchmark of 45 trajectory inference methods}
	\begin{center}
		\only<1>{\includegraphics[width=.9\linewidth]{figures/5a_benchmark.pdf}}%
		\only<2>{\includegraphics[width=.9\linewidth]{figures/5b_benchmark.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Interactive guidelines}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/6_guidelines.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{SCORPIUS: Linear trajectory inference}
	\begin{center}
		\includegraphics[width=.8\linewidth]{figures/end_part_scorpius.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{dyno: A toolkit for inferring and interpreting trajectories}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/end_part_dyno.pdf}
	\end{center}
\end{frame}
\begin{frame}
	\frametitle{Perspective: self-assessment in trajectory inference}
	\begin{center}
		\includegraphics[width=.8\linewidth]{figures/end_part_selfassessment.pdf}
	\end{center}
\end{frame}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{lightgray}{black}{lightgray}{lightgray}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{lightgray}{lightgray}{black}{lightgray}

\begin{frame}
	\frametitle{dyngen}
	\begin{center}
		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
	\end{center}
\end{frame}

\begin{frame}
\frametitle{Guidelines for benchmarking computational tools}
\begin{center}
	\includegraphics[width=\linewidth]{figures/end_part_guidelines.pdf}
\end{center}
\end{frame}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{lightgray}{lightgray}{lightgray}{black}

>>>>

\newcommand{\summary}[5]{%
	\begin{frame}
		\begin{itemize}
			\item {\color{#1}Single-cell biology}
			\item {\color{#2}Trajectory inference}
			\item {\color{#3}Network inference}
			\item {\color{#4}Benchmarking}
			\item {\color{#5}Practical implications}
		\end{itemize}
	\end{frame}
}
\newcommand{\topright}[1]{%
\begin{textblock*}{.15\linewidth}(.98\linewidth,.05\linewidth)
	\includegraphics[width=\linewidth]{figures/overview_#1.pdf}
\end{textblock*}}

@@

%%%%%%%%%%%%%% SINGLE CELL BIOLOGY %%%%%%%%%%%%%%
\begin{frame}
	\frametitle{Single-cell biology \uncover<2->{is to study an organism}\\\uncover<3->{by studying the behaviour of its cells}}
	\begin{center}
		\only<2>{\includegraphics[width=\linewidth]{figures/1a_singlecellbiology}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/1b_singlecellbiology}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/1c_singlecellbiology}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Technological advances allow high-throughput\\single-cell biology}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/2_part_scaling.pdf}\\
		\flushright 		{\vspace*{-1em}\footnotesize Regev et al. 2018}
	\end{center}
  \vspace*{-2.2em}
\end{frame}

\begin{frame}
	\frametitle{Research questions}
	\begin{center}
		\only<2>{\includegraphics[width=\linewidth]{figures/11a_fundamental.pdf}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/11b_fundamental.pdf}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/11c_fundamental.pdf}}%
		\only<5>{\includegraphics[width=\linewidth]{figures/11d_fundamental.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
\includegraphics[width=\linewidth]{figures/overview.pdf}
\end{frame}

%%%%%%%%%%%%%% TRAJECTORY INFERENCE %%%%%%%%%%%%%%

\begin{frame}
	\vfill
	\begin{center}
		\LARGE Trajectory inference
	\end{center}
	\vfill
\end{frame}

@@

\begin{frame}
	\frametitle{dyngen: Benchmarking with \textit{in silico} cells}
	\begin{center}
		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
	\end{center}
	\topright{ch2}
\end{frame}

\begin{frame}
	\frametitle{Benchmark of 45 trajectory inference methods}
%	\begin{center}
		\only<1>{\includegraphics[width=.9\linewidth]{figures/5a_benchmark.pdf}}%
		\only<2>{\includegraphics[width=.9\linewidth]{figures/5b_benchmark.pdf}}%
%	\end{center}
	\topright{ch3}
\end{frame}

\begin{frame}
	\frametitle{Interactive guidelines}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/6_guidelines.pdf}
	\end{center}
	\topright{ch3}
\end{frame}

\begin{frame}
	\frametitle{Perspective: self-assessment in trajectory inference}
	\begin{center}
		\includegraphics[width=.8\linewidth]{figures/end_part_selfassessment.pdf}
	\end{center}
  \topright{ch8}
\end{frame}

\begin{frame}
	\frametitle{SCORPIUS: Linear trajectory inference}
	\begin{center}
		\includegraphics[width=.8\linewidth]{figures/end_part_scorpius.pdf}
	\end{center}
	\topright{ch5}
\end{frame}

\begin{frame}
	\frametitle{dyno: A toolkit for inferring and\\interpreting trajectories}
	\begin{center}
		\includegraphics[width=.9\linewidth]{figures/end_part_dyno.pdf}
	\end{center}
	\topright{ch4}
\end{frame}

%%%%%%%%%%%%%% NETWORK INFERENCE %%%%%%%%%%%%%%

\begin{frame}
	\vfill
	\begin{center}
		\LARGE Network inference
	\end{center}
	\vfill
\end{frame}

\begin{frame}
	\frametitle{Cellular dynamics is (i.a.) driven by gene regulation}
	\begin{center}
		\only<1>{\includegraphics[width=.8\linewidth]{figures/7a_generegulation.pdf}}%
		\only<2>{\includegraphics[width=.8\linewidth]{figures/7b_generegulation.pdf}}%
		\only<3>{\includegraphics[width=.8\linewidth]{figures/7c_generegulation.pdf}}%
		\only<4>{\includegraphics[width=.8\linewidth]{figures/7d_generegulation.pdf}}%
		\only<5>{\includegraphics[width=.8\linewidth]{figures/7e_generegulation.pdf}}%
		\only<6>{\includegraphics[width=.8\linewidth]{figures/7f_generegulation.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Classical network inference}
	\begin{center}
		\includegraphics[width=.9\linewidth]{figures/8_networkinference.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Case-specific network inference}
	\begin{center}
		\only<1>{\includegraphics[width=\linewidth]{figures/9a_csni_example.pdf}}%
		\only<2>{\includegraphics[width=\linewidth]{figures/9b_csni_example.pdf}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/9c_csni_example.pdf}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/9d_csni_example.pdf}}%
	\end{center}
	\topright{ch6}
\end{frame}

\begin{frame}
	\frametitle{bred applied to TCGA}
	\begin{center}
		\includegraphics[width=.85\linewidth]{figures/10b_tcga.pdf}
%		\only<1>{\includegraphics[width=.6\linewidth]{figures/10a_tcga.pdf}}%
%		\only<2>{\includegraphics[width=.85\linewidth]{figures/10b_tcga.pdf}}%
	\end{center}
	\topright{ch6}
\end{frame}

\begin{frame}
	\frametitle{incgraph: Optimising regulatory networks}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/end_part_incgraph.pdf}
	\end{center}
	\topright{ch7}
\end{frame}

%%%%%%%%%%%%%%%%%%% BENCHMARKING
\begin{frame}
	\vfill
	\begin{center}
		\LARGE Benchmarking
	\end{center}
	\vfill
\end{frame}

\begin{frame}
	\frametitle{Essential guidelines for benchmarking\\computational tools}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/end_part_guidelines.pdf}
	\end{center}
	\topright{ch9}
\end{frame}

%%%%%%%%%%%%%% IMPLICATIONS %%%%%%%%%%%%%%

\begin{frame}
	\frametitle{Practical implications}
	\begin{itemize}
		\item<2-> Fundamental research in single-cell analyses \\
		  \uncover<3->{\begin{center}\includegraphics[width=.7\linewidth]{figures/11d_fundamental.pdf}\end{center}}
		\item<4-> Benchmarking
		\item<5-> R \& tidyverse
	\end{itemize}
  \topright{all}
\end{frame}
